\emph{On Logics Extended with Embedding-Closed Quantifiers}

### 2020-11-06 paper.tex
Generalize finite case chain results to interpretations

\begin{lemma}\label{quant_elim_finite_lemma}
Let $\tau$ be a finite relational vocabulary, $Q$ an embedding-closed quantifier
of width $n<\omega$ and $\varphi = Q(x_i\psi_i)_{i<n}$ where all $\psi_i$ are
quantifier-free $\tau$-formulas.
Let $(\mathfrak{A}_i)_{i<\omega}$ be a chain of quasi-homogeneous $\tau$-structures.
Then there is a natural number $k$ and a quantifier-free $\tau$-formula
$\vartheta$ such that
\[
	\mathfrak{A}_i \vDash \forall \overline{x}(\varphi \leftrightarrow \vartheta)
\]
for all $i \geq k$.
\end{lemma}
\begin{proof}
For each $i < \omega$, let
\[
	T_i = \{t \colon t \text{ is an atomic type and } (\mathfrak{A}_i,\overline{a}) \vDash \varphi \wedge  t \text{ for some } \overline{a} \}.
\]
Since all $\mathfrak{A}_i$ are quasi-homogeneous, it follows from Theorem
\ref{homog} and Lemma \ref{atomic_types} that
\[
	\mathfrak{A}_i \vDash \forall \overline{x}(\varphi \leftrightarrow \bigvee T_i).
\]
Now let $i \leq j < \omega$ and $t \in T_i$. We have
$(\mathfrak{A}_i,\overline{a}) \vDash \varphi \wedge t$ for some
$\overline{a}$, so $(\mathfrak{A}_j,\overline{a}) \vDash \varphi \wedge t$ since
both $\varphi$ and $t$ are preserved by embeddings.
Thus, $t \in T_j$, so $T_i \subseteq T_j$ always when $i \leq j$.
Since there are finitely many distinct atomic $n$-types because $\tau$ is finite
and relational, the chain $(T_i)_{i < \omega}$ reaches its maximum at some
$k < \omega$. Then $\vartheta = \bigvee T_k$ is a quantifier-free $\tau$-formula
we want.
\end{proof}

\begin{theorem}\label{finite}
Let $\tau$ be a finite relational vocabulary, $\mathcal{Q}$ a finite set of
embed\-ding-closed quantifiers of finite width and
$(\mathfrak{A}_i)_{i<\omega}$ a chain of quasi-ho\-mo\-ge\-neous $\tau$-structures.
For each $m< \omega$, there is a natural number $N_m$ such that for every formula
$\varphi \in \mathcal{L}^m_{\infty \omega}(\mathcal{Q})[\tau]$ there is a
quantifier-free $\tau$-formula $\vartheta_{\varphi}$ such that
\[
	\mathfrak{A}_i \vDash \forall \overline{x}(\varphi \leftrightarrow \vartheta_{\varphi})
\]
for all $i \geq N_m$.
Note that $N_m$ depends on the chain of $\tau$-structures.
\end{theorem}
\begin{proof}
Let $\psi_0,\dots,\psi_l$ be an enumeration of all (up to equivalence) the
$\tau$-formulas in at most $m$ variables having form
$Q(\overline{x}_i\vartheta_i)_{i<n}$ with $Q \in \mathcal{Q}$ and all
$\vartheta_i$ quantifier-free.
Note that $l$ is finite.
By Lemma \ref{quant_elim_finite_lemma} for each $\psi_i$ there is
$k_i < \omega$ and a quantifier-free $\tau$-formula $\vartheta$ such that
\[
	\mathfrak{A}_j \vDash \forall \overline{x} (\psi_i \leftrightarrow \vartheta)
\]
when $j \geq k_i$.
We claim that we can set $N_m := \max\{k_i : i \leq l\}$.
We prove the claim by induction on the structure of the formula $\varphi$.
The cases of $\varphi$ atomic, $\varphi = \neg \alpha$ and
$\varphi = \bigwedge_{i \in I}\alpha_i$ are clear.
Suppose $\varphi = Q(\overline{x}_i\alpha_i)_{i<n}$ and the claim holds for all
$\alpha_i$.
Then for each $i < n$ there is a quantifier-free $\tau$-formula $\vartheta_i$
such that
\[
	\mathfrak{A}_j \vDash \forall \overline{y}(\alpha_i \leftrightarrow \vartheta_i)
\]
for all $j \geq N_m$ so
\[
	\mathfrak{A}_j \vDash \forall \overline{y}(\varphi \leftrightarrow Q(\overline{x}_i\vartheta_i)_{i<n})
\]
when $j \geq N_m$. Thus, if $j \geq N_m$, then $\varphi$ is equivalent to some
$\psi_r$ and therefore to some quantifier-free formula $\vartheta$, so we can
set $\vartheta_{\varphi} := \vartheta$.
\end{proof}

@@

\begin{corollary}\label{finite2}
Let $\tau$ be a finite relational vocabulary, $\mathcal{Q}$ a finite set of
embed\-ding-closed quantifiers of finite width and
$(\mathfrak{A}_i)_{i<\omega}$ a chain of quasi-ho\-mo\-ge\-neous $\tau$-structures.
Let $m<\omega$ and
write $\mathcal{L} = \mathcal{L}_{\infty\omega}^m(\mathcal{Q})$.
There is a natural number $N_m$ such that
$\mathfrak{A}_i \preceq_{\mathcal{L}} \mathfrak{A}_j$ for all $j \geq i \geq N_m$.
\end{corollary}

@@

As we saw in Example \ref{hom_graphs}, there are only few homogeneous finite
graphs, so we cannot usually directly apply Corollary \ref{finite2} in studying
definability of graph properties.
The following theorem shows that this situation can be remedied to some extent
by using \emph{interpretations}.

\begin{theorem}\label{inter}
Let $\tau$ and $\sigma$ be finite relational vocabularies and suppose
$\mathcal{Q}$ is a finite set of embedding-closed quantifiers of finite width.
Let $m < \omega$, and write
$\mathcal{L} = \mathcal{L}^m_{\infty\omega}(\mathcal{Q})$.
Suppose $(\mathfrak{A}_i)_{i<\omega}$ is a chain of quasi-homogeneous
$\tau$-structures and $(\Psi,(\psi_R)_{R\in\sigma})$ is an
$\mathcal{L}$-interpretation of $\sigma$ in $\tau$.
There is a natural number $N_m$ such that for all $j \geq i \geq N_m$ we have
$\Psi(\mathfrak{A}_i) \preceq_{\mathcal{L}} \Psi(\mathfrak{A}_j)$.
\end{theorem}
\begin{proof}
By Corollary \ref{finite2}, there is a natural number $N'_m$ such that
$\mathfrak{A}_i \preceq_{\mathcal{L}} \mathfrak{A}_j$ for all $j\geq i \geq N'_m$.
We claim that $N'_m$ is a wanted number $N_m$.
To show that, let $\varphi \in \mathcal{L}[\sigma]$.
By Lemma \ref{interpr}, there is a formula $\varphi^* \in \mathcal{L}[\tau]$ such that
\[
	\mathfrak{A}_i, \overline{a} \vDash \varphi^* \Leftrightarrow \Psi(\mathfrak{A}_i), \overline{a} \vDash \varphi
\]
for all $i<\omega$ and tuples $\overline{a}$ of elements of $A$.
Let $j \geq i \geq N'_m$ and suppose that
$\Psi(\mathfrak{A}_i),\overline{a} \vDash \varphi$.
Then we have
$\mathfrak{A}_i,\overline{a} \vDash \varphi^*$,
so $\mathfrak{A}_j, \overline{a} \vDash \varphi^*$ from which follows
$\Psi(\mathfrak{A}_j), \overline{a} \vDash \varphi$.
The claim is thus proved.
\end{proof}

\begin{example}
A graph $G$ is \emph{regular} if every vertex of $G$ has the same number of neighbours.
We will show that regularity of a graph is not definable in the logic
$\mathcal{L} = \mathcal{L}^{\omega}_{\infty \omega}(Q_1,\dots,Q_n)$ where all
$Q_i \in \mathcal{Q}_{\emb}$ have finite width.
Let $\tau$ be the same vocabulary and $(\mathfrak{A}_i)_{i<\omega}$ a chain of
$\tau$-structures as in Example \ref{Hartig_ex}.
Let $\sigma = \{E\}$ where $E$ is a relation symbol and suppose
$(\Psi,\psi)$ is an interpretation of $\sigma$ in $\tau$ where
\[
	\psi(x,y) := (U(x)\wedge U(y)) \vee (V(x) \wedge V(y)).
\]
Then for all $k<\omega$,
\[
	\Psi(\mathfrak{A}_k) = \begin{cases}
	K_{\frac{k+1}{2}} + K_{\frac{k+1}{2}} \text{ if } k \text{ is odd} \\
	K_{\frac{k}{2}} + K_{\frac{k}{2}+1} \text{ if } k \text{ is even},
	\end{cases}
\]
where every $K_m$ is the complete graph on $m$ vertices and $+$ means disjoint
union (cf. Example \ref{hom_graphs}).
Thus, a graph $\Psi(\mathfrak{A}_k)$ is regular if and only if $k$ is odd,
so by Theorem \ref{inter} regularity of graphs is not definable in $\mathcal{L}$.
\end{example}

>>>>

\begin{lemma}\label{quant_elim_finite_lemma}
Let $\tau$ and $\sigma$ be finite relational vocabularies and
$(\Psi, (\psi_R)_{R \in \sigma})$ an
$\mathcal{L}_{\infty \omega}(\mathcal{Q}_{\emb})$-interpretation of
$\sigma$ in $\tau$.
Suppose that $Q$ is an embedding-closed quantifier of width $n < \omega$
and $\varphi = Q(\overline{x}_i \chi_i)_{i < n}$ a formula with $m$ free
variables where all $\chi_i$ are
quantifier-free $\sigma$-formulas.
Let $(\Psi(\mathfrak{A}_i))_{i < \omega}$ be a chain where all $\mathfrak{A}_i$
are quasi-homogeneous $\tau$-structures.
Then there is a natural number $N$ and a quantifier-free $\sigma$-formula
$\vartheta$ such that
\[
	\Psi(\mathfrak{A}_i) \vDash \forall
	\overline{x}(\varphi \leftrightarrow \vartheta)
\]
for all $i \geq N$.
\end{lemma}
\begin{proof}
For each $i < \omega$, let $T_i$ be the set of all atomic
$m$-types $t$ of $\tau$
such that $\mathfrak{A}_j, \overline{a} \vDash t$
and $\Psi(\mathfrak{A}_j), \overline{a} \vDash \varphi$
for some $\overline{a} \in A^m$ and $j \leq i$.
We clearly have $T_j \subseteq T_i$ whenever $j \leq i$.
Since $\tau$ is finite and relational, there is a natural number $k$
at which the chain $(T_i)_{i < \omega}$ achieves its maximum.
Write $T = T_k$.
We claim that for all $i \geq k$ and $\overline{a} \in A_i^m$, we have
\begin{equation}\label{eq_0}
\mathfrak{A}_i, \overline{a} \vDash \bigvee T
\Leftrightarrow
\Psi(\mathfrak{A}_i), \overline{a} \vDash \varphi.
\end{equation}
The statement on the right side implies the left side statement by the
definition of $T$.
For the other direction, suppose that
$\mathfrak{A}_i, \overline{a} \vDash \bigvee T$
and assume to the contrary that
$\Psi(\mathfrak{A}_i), \overline{a} \nvDash \varphi$.
By the definition of $T$, there is $j \leq i$ and $\overline{b} \in A_j^m$
such that $\overline{a}$ and $\overline{b}$ have the same atomic type of $\tau$
and $\Psi(\mathfrak{A}_j), \overline{b} \vDash \varphi$.
Since $\Psi(\mathfrak{A}_j) \leq \Psi(\mathfrak{A}_i)$, $Q$ is embedding-closed
and all $\chi_i$ are quantifier-free, we have
$\Psi(\mathfrak{A}_i), \overline{b} \vDash \varphi$.
However, since $\mathfrak{A}_i$ is quasi-homogeneous,
there is, by Lemma 2.2 and Theorem 5.3, a quantifier-free $\tau$-formula
$\varphi^*$ such that
$\mathfrak{A}_i, \overline{a} \nvDash \varphi^*$
and $\mathfrak{A}_i, \overline{b} \vDash \varphi^*$
which is a contradiction since $\overline{a}$ and $\overline{b}$ have the
same atomic types in $\tau$.
The claim is thus proven.

For each natural number $i \geq k$,
let $T'_i$ be the set of all atomic types $t$ of $\sigma$
for which there is a natural number $j$ with
$k \leq j \leq i$ and $\overline{a} \in A^m_j$ such that
$\Psi(\mathfrak{A}_j), \overline{a} \vDash t$ and
$\mathfrak{A}_j, \overline{a} \vDash \bigvee T$.
The chain $(T'_i)_{i < \omega}$ is clearly monotone and, since $\sigma$ is
finite and relational, there is a natural
number $k'$ at which the chain $(T'_i)_{i < \omega}$ reaches
its maximum.
Denote $T' = T'_{k'}$.
We claim that
\begin{equation}\label{eq_1}
\Psi(\mathfrak{A}_i), \overline{a} \vDash \bigvee T' \Leftrightarrow
\mathfrak{A}_i, \overline{a} \vDash \bigvee T
\end{equation}
for all $i \geq k'$ and $\overline{a} \in A_i^m$.

"$\Rightarrow$" Suppose we have
$\Psi(\mathfrak{A}_i), \overline{a} \vDash \bigvee T'$ for some $i \geq k'$
and $\overline{a} \in A_i^m$.
Then for some $j$ with $k \leq j \leq k'$ there is $\overline{a} \in A_j^m$ such that
$\mathfrak{A}_j, \overline{a} \vDash \bigvee T$.
By (\ref{eq_0}) it follows that $\Psi(\mathfrak{A}_j), \overline{a} \vDash \varphi$
and, since $\Psi(\mathfrak{A}_j) \leq \Psi(\mathfrak{A}_i)$, we have
$\Psi(\mathfrak{A}_i), \overline{a} \vDash \varphi$ and thus
$\mathfrak{A}_i, \overline{a} \vDash \bigvee T$.

"$\Leftarrow$" Let $\mathfrak{A}_i, \overline{a} \vDash \bigvee T$
for some $i \geq k'$ and $\overline{a} \in A_i^m$.
Assume to the contrary that
$\Psi(\mathfrak{A}_i), \overline{a} \nvDash \bigvee T'$.
Then the atomic type $t$ of $\sigma$ for which
$\Psi(\mathfrak{A}_i), \overline{a} \vDash t$ is in $T_i$ but not in $T_k'$
which contradicts the set's $T_k'$ maximality.
The claim (\ref{eq_1}) thus holds.

Thus, by combining the claims (\ref{eq_0}) and (\ref{eq_1}) we get
\[
	\Psi(\mathfrak{A}_i), \overline{a} \vDash \bigvee T' \Leftrightarrow
	\Psi(\mathfrak{A}_i), \overline{a} \vDash \varphi
\]
for all $i \geq k'$ and $\overline{a} \in A_i^k$ so we can set
$N := k'$ and $\vartheta := \bigvee T'$ to get the desired result.
\end{proof}

\begin{theorem}\label{finite}
Let $\tau$ and $\sigma$ be finite relational vocabularies,
$(\Psi, (\psi_R)_{R \in \sigma})$ an
$\mathcal{L}_{\infty \omega}(\mathcal{Q}_{\emb})$-interpretation of
$\sigma$ in $\tau$, and $(\Psi(\mathfrak{A}_i))_{i < \omega}$
a chain in which all $\mathfrak{A}_i$ are quasi-homogeneous $\tau$-structures.
Let $\mathcal{Q}$ be a finite set of embedding-closed quantifiers of finite width.
For each $m< \omega$, there is a natural number $N_m$ such that for every formula
$\varphi \in \mathcal{L}^m_{\infty \omega}(\mathcal{Q})[\sigma]$ there is a
quantifier-free $\tau$-formula $\vartheta_{\varphi}$ such that
\[
	\Psi(\mathfrak{A}_i) \vDash \forall \overline{x}(\varphi
	\leftrightarrow \vartheta_{\varphi})
\]
for all $i \geq N_m$.
Note that $N_m$ depends on the chain of structures.
\end{theorem}
\begin{proof}
Let $\chi_0,\dots,\chi_l$ be an enumeration of all (up to equivalence) the
$\sigma$-formulas in at most $m$ variables having form
$Q(\overline{x}_i\vartheta_i)_{i<n}$ with $Q \in \mathcal{Q}$ and all
$\vartheta_i$ quantifier-free.
Note that $l$ is finite.
By Lemma \ref{quant_elim_finite_lemma} for each $\chi_i$ there is
$k_i < \omega$ and a quantifier-free $\sigma$-formula $\vartheta$ such that
\[
	\Psi(\mathfrak{A}_j) \vDash \forall \overline{x} (\chi_i \leftrightarrow \vartheta)
\]
when $j \geq k_i$.
We claim that we can set $N_m := \max\{k_i : i \leq l\}$.
We prove the claim by induction on the structure of the formula $\varphi$.
The cases of $\varphi$ atomic, $\varphi = \neg \alpha$ and
$\varphi = \bigwedge_{i \in I}\alpha_i$ are clear.
Suppose $\varphi = Q(\overline{x}_i\alpha_i)_{i<n}$ and the claim holds for all
$\alpha_i$.
Then for each $i < n$ there is a quantifier-free $\sigma$-formula $\vartheta_i$
such that
\[
	\Psi(\mathfrak{A}_j) \vDash \forall \overline{y}(\alpha_i
	\leftrightarrow \vartheta_i)
\]
for all $j \geq N_m$ so
\[
	\Psi(\mathfrak{A}_j) \vDash \forall \overline{y}(\varphi
	\leftrightarrow Q(\overline{x}_i\vartheta_i)_{i<n})
\]
when $j \geq N_m$. Thus, if $j \geq N_m$, then $\varphi$ is equivalent to some
$\chi_r$ and therefore to some quantifier-free formula $\vartheta$, so we can
set $\vartheta_{\varphi} := \vartheta$.
\end{proof}

@@

\begin{corollary}\label{finite2}
Let $\tau$ and $\sigma$ be finite relational vocabularies and suppose
$\mathcal{Q}$ is a finite set of embedding-closed quantifiers of finite width.
Let $m < \omega$, and write
$\mathcal{L} = \mathcal{L}^m_{\infty\omega}(\mathcal{Q})$.
Suppose $(\mathfrak{A}_i)_{i<\omega}$ is a chain of quasi-homogeneous
$\tau$-structures and $(\Psi,(\psi_R)_{R\in\sigma})$ is an
$\mathcal{L}$-interpretation of $\sigma$ in $\tau$.
There is a natural number $N_m$ such that for all $j \geq i \geq N_m$ we have
$\Psi(\mathfrak{A}_i) \preceq_{\mathcal{L}} \Psi(\mathfrak{A}_j)$.
\end{corollary}

@@

\begin{example}\label{reg_graph_ex}
A graph $G$ is \emph{regular} if every vertex of $G$ has the same number of neighbours.
We will show that regularity of a graph is not definable in the logic
$\mathcal{L} = \mathcal{L}^{\omega}_{\infty \omega}(Q_1,\dots,Q_n)$ where all
$Q_i \in \mathcal{Q}_{\emb}$ have finite width.
Let $\tau$ be the same vocabulary and $(\mathfrak{A}_i)_{i<\omega}$ a chain of
$\tau$-structures as in Example \ref{Hartig_ex}.
Let $\sigma = \{E\}$ where $E$ is a binary relation symbol and suppose
$(\Psi,\psi)$ is an interpretation of $\sigma$ in $\tau$ where
\[
	\psi(x,y) := (U(x)\wedge U(y)) \vee (V(x) \wedge V(y)).
\]
Then for all $k<\omega$,
\[
	\Psi(\mathfrak{A}_k) = \begin{cases}
	K_{\frac{k+1}{2}} + K_{\frac{k+1}{2}} \text{ if } k \text{ is odd} \\
	K_{\frac{k}{2}} + K_{\frac{k}{2}+1} \text{ if } k \text{ is even},
	\end{cases}
\]
where every $K_m$ is the complete graph on $m$ vertices and $+$ means disjoint
union (cf. Example \ref{hom_graphs}).
Thus, a graph $\Psi(\mathfrak{A}_k)$ is regular if and only if $k$ is odd,
so by Corollary \ref{finite2} regularity of graphs is not definable in $\mathcal{L}$.
\end{example}
